---
title: "Final Project - EDLD 652"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    source: http://dwfainstein.shinyapps.io/Final_proj_edld652
---

```{r setup, include=FALSE}
library(flexdashboard)
library(here)
library(readxl)
library(mapcan)
library(tidyverse)
library(janitor)
library(shiny)
library(plotly)
library(stringr)
library(viridis)
library(ggpubr)
library(RColorBrewer)
```

``` {r prep}
composition_data <- read_excel("data/class_size_composition.xlsx") %>% 
  filter(DATA_LEVEL == "DISTRICT LEVEL",
         GRADE_GROUP == "All Grades")

districting <- read_excel("data/DataVizfinal_names.xlsx") %>% 
  select(SD,
         DISTRICT_NAME,
         riding_name_english,
         Code) %>% 
  full_join(composition_data, by = "DISTRICT_NAME") %>% 
  rename(riding_code = Code)

districting$SCHOOL_YEAR <- str_sub(districting$SCHOOL_YEAR, 6)

```

BC Education Data {.sidebar}
===================================
This project portrays 2020 geographic, and 15-year longitudinal data on English Learners and students with exceptionalities in the province of British Columbia. Navigate to multiple interactive figures that indicate the amount of **classrooms** in each district that are composed of at least three students with individualized education plans or English Learners. 

*Note: the interactive maps include federal riding regions, not school district regions. Therefore, values are mapped approximately to locale and some locations have insufficient data.*

All data is gathered from the BC Minsitry of Education [open-source data files](https://catalogue.data.gov.bc.ca/dataset). This is an open-source data science project for EDLD 652 at the University of Oregon produced by [David Fainstein](https://twitter.com/TeachAbilities).


Individualized Education Plans
===============================
Column {.tabset}
----------------------
### Classes with Minimum Three Students with IEPs

```{r setMap}
BC_data <- mapcan::census_pop2016 %>% 
  select(population_2016,
         pr_english,
         population_density_2016,
         born_outside_canada,
         census_division_name) %>% 
  filter(pr_english == "British Columbia") %>% 
    mapcan(boundaries = ridings,
          type = standard,
          province = BC)

IEPex1 <- districting %>% 
  filter(SCHOOL_YEAR == "2020") %>% 
  select( NUMBER_CLASSES_3_IEP,
          NUMBER_CLASSES_4_IEP,
          NUMBER_CLASSES_5_IEP,
          NUMBER_CLASSES_6_IEP,
          NUMBER_CLASSES_7_PLUS_IEP,
         riding_code) %>% 
  group_by(riding_code) %>% 
  summarise(`Classes >2 IEPs` = sum(NUMBER_CLASSES_3_IEP,
                            NUMBER_CLASSES_4_IEP,
                            NUMBER_CLASSES_5_IEP,
                            NUMBER_CLASSES_6_IEP,
                            NUMBER_CLASSES_7_PLUS_IEP)) %>%
  drop_na() %>% 
  full_join(BC_data, by = "riding_code") %>% 
  rename("Region" = riding_name_english) 

IEP <- ggplot(IEPex1, aes(x = long, y = lat, group = group, fill = `Classes >2 IEPs`, A = Region, B = `Classes >2 IEPs`)) +
  geom_polygon() +
  coord_fixed() +
  theme_mapcan() +
  scale_fill_viridis_c(name = "Total Classes with at least 3 \n Students with IEPs") #+
  # theme(legend.box = "vertical")

ggplotly(IEP, tooltip = c("A", "B"))
# ggplotly(EL)
# gridExtra::grid.arrange(IEP, EL, ncol = 2)
# plotly::ggplotly(IEP)
# renderPlotly(IEP)

```
### Longitudinal View

```{r IEP_time}

Plot_Longitudinal1 <- districting %>% 
  mutate(SCHOOL_YEAR = as.numeric(SCHOOL_YEAR),
          `+3 IEP Classrooms` = NUMBER_CLASSES_3_IEP,
          NUMBER_CLASSES_4_IEP,
          NUMBER_CLASSES_5_IEP,
          NUMBER_CLASSES_6_IEP,
          NUMBER_CLASSES_7_PLUS_IEP,) %>% 
  rename("Year" = SCHOOL_YEAR,
         "District" = DISTRICT_NAME) 

Plot_Longitudinal1 <- highlight_key(Plot_Longitudinal1, ~District)

longitudinalBC1 <- ggplot(Plot_Longitudinal1, aes(Year, `+3 IEP Classrooms`))+
  geom_line(aes(group = District),
            # color = "gray75",
            alpha = 0.3) +
  # geom_line(data = filter(Plot_Longitudinal1,
  #     DISTRICT_NAME == "Surrey"),
  #          color = "#D95F02") +
  # geom_line(data = filter(Plot_Longitudinal1,
  #     DISTRICT_NAME == "Vancouver"),
  #     color = "#7570B3") +
  # geom_line(data = filter(Plot_Longitudinal1,
  #     DISTRICT_NAME == "Greater Victoria"),
  #          color = "#E7298A") +
  theme_minimal() +
  labs(title = "Classrooms with at least 3 IEPs by School District") +
       xlab("School Year") +
       ylab("Number of Classrooms with >= 3 IEPs") +
  theme(legend.position = "none")
    # annotate(geom = "text", x = 2012, y = 830, label = "Surrey", hjust = "left") +
    # annotate(geom = "text", x = 2008, y = 1080, label = "Vancouver", hjust = "left") +
    # annotate(geom = "text", x = 2012, y = 340, label = "Victoria", hjust = "left")

cols <- toRGB(brewer.pal(3, "Dark2"), 0.5)
IEP_long <- ggplotly(longitudinalBC1, tooltip = c("District", "Year", "+3 IEP Classrooms"))
highlight(IEP_long, on = "plotly_hover", color = cols, dynamic = TRUE)

```



English Learners
=============================

Column {.tabset}
-----------------------------------------------------------------------

### Classes with Minimum Three Students Designated as English Learners

```{r ELexample}

ELex1 <- districting %>% 
  filter(SCHOOL_YEAR == "2020") %>% 
  select(NUMBER_CLASSES_1_ELL,
         NUMBER_CLASSES_2_ELL,
         NUMBER_CLASSES_3_ELL,
         NUMBER_CLASSES_4_ELL,
         NUMBER_CLASSES_5_ELL,
         NUMBER_CLASSES_6_ELL,
         NUMBER_CLASSES_7_PLUS_ELL,
         riding_code) %>%
  group_by(riding_code) %>%
  summarise(`Classes >2 ELs` = sum( NUMBER_CLASSES_1_ELL, 
                            NUMBER_CLASSES_2_ELL,
                            NUMBER_CLASSES_3_ELL,
                            NUMBER_CLASSES_4_ELL,
                            NUMBER_CLASSES_5_ELL,
                            NUMBER_CLASSES_6_ELL,
                            NUMBER_CLASSES_7_PLUS_ELL)) %>%
  drop_na() %>%
  full_join(BC_data, by = "riding_code") %>% 
  rename("Region" = riding_name_english)

EL <- ggplot(ELex1, aes(x = long, y = lat, group = group, fill = `Classes >2 ELs`, A = Region, B = `Classes >2 ELs`)) +
  geom_polygon() +
  coord_fixed() +
  theme_mapcan() +
  scale_fill_viridis_c(name = "Total Classes with \n at least 3 ELs")
  

ggplotly(EL, tooltip = c("A", "B"))

```

### Longitudinal View


```{r EL_time}

Plot_Longitudinal <- districting %>% 
  mutate(SCHOOL_YEAR = as.numeric(SCHOOL_YEAR),
          `Classrooms with +3 ELs` = NUMBER_CLASSES_3_ELL +
           NUMBER_CLASSES_4_ELL +
           NUMBER_CLASSES_5_ELL +
           NUMBER_CLASSES_6_ELL +
           NUMBER_CLASSES_7_PLUS_ELL) %>% 
  rename("Year" = SCHOOL_YEAR,
         "District" = DISTRICT_NAME) 

Plot_Longitudinal <- highlight_key(Plot_Longitudinal, ~District)

longitudinalBC <- ggplot(Plot_Longitudinal, aes(Year, `Classrooms with +3 ELs`), District, Year, `Classrooms with +3 ELs`)+
  geom_line(aes(group = District),
            alpha = 0.3) +
  # geom_line(data = filter(Plot_Longitudinal,
  #     DISTRICT_NAME == "Surrey"),
  #          color = "#D95F02") +
  # geom_line(data = filter(Plot_Longitudinal,
  #     DISTRICT_NAME == "Vancouver"),
  #     color = "#7570B3") +
  # geom_line(data = filter(Plot_Longitudinal,
  #     DISTRICT_NAME == "Greater Victoria"),
  #          color = "#E7298A") +
  theme_minimal() +
  labs(title = "Classrooms with at least 3 ELs by School District") +
       xlab("School Year") +
       ylab("Number of Classrooms with >= 3 ELs") +
  theme(legend.position = "none") 
    # annotate(geom = "text", x = 2017, y = 3250, label = "Surrey", hjust = "left") +
    # annotate(geom = "text", x = 2008, y = 3350, label = "Vancouver", hjust = "left") +
    # annotate(geom = "text", x = 2012, y = 500, label = "Victoria", hjust = "left")

# cols <- toRGB(brewer.pal(3, "Dark2"), 0.5)

p <- ggplotly(longitudinalBC, tooltip = c("District", "Year", "Classrooms with +3 ELs"))
highlight(p, on = "plotly_hover", color = cols, dynamic = TRUE)

```

Combined Map
=============================

Column {data-width=350}
-----------------------------------------------------------------------

### EL Juxtaposed IEP


```{r Combined}

library(patchwork)

IEP + EL + 
  plot_layout(guides = "collect") & 
  theme(legend.position = "bottom",
        legend.title = element_text(size = 10),
        legend.text = element_text(size = 6)) 

````



