---
title: "Final Project - EDLD 652"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    source: https://github.com/dwfainstein/Final_proj_edld652
---

```{r setup, include=FALSE}
library(flexdashboard)
library(here)
library(readxl)
library(mapcan)
library(tidyverse)
library(janitor)
library(shiny)
library(plotly)
library(stringr)
```

``` {r prep}
composition_data <- read_excel(here("data", "class_size_composition.xlsx")) %>% 
  filter(DATA_LEVEL == "DISTRICT LEVEL",
         GRADE_GROUP == "All Grades")

districting <- read_excel(here("data", "DataVizfinal_names.xlsx")) %>% 
  select(SD,
         DISTRICT_NAME,
         riding_name_english,
         Code) %>% 
  full_join(composition_data, by = "DISTRICT_NAME") %>% 
  rename(riding_code = Code)

districting$SCHOOL_YEAR <- str_sub(districting$SCHOOL_YEAR, 6)

```

BC Education Data {.sidebar}
===================================
This project portrays 2020 geographic, and 15-year longitudinal data on English Learner and students with exceptionalities in the province of British Columbia. Navigate to multiple interactive figures that indicate the amount of **classrooms** in each district that are composed of at least three students with individualized education plans or English Learners. Three students with IEPs **or** identified as ELs is the codified threshold that indicates teachers **may** be best supported with additional resources for providing effective instruction to all learners in their classroom.

All data is gathered from the BC Minsitry of Education [open-source data files](https://catalogue.data.gov.bc.ca/dataset). This is an open-source data science project for EDLD 652 at the University of Oregon produced by [David Fainstein](https://twitter.com/TeachAbilities).


Individualized Education Plans
===============================
Column {.tabset}
----------------------
### Classes with Minimum Three Students with IEPs

```{r setMap}
BC_data <- mapcan::census_pop2016 %>% 
  select(population_2016,
         pr_english,
         population_density_2016,
         born_outside_canada,
         census_division_name) %>% 
  filter(pr_english == "British Columbia") %>% 
    mapcan(boundaries = ridings,
          type = standard,
          province = BC)

IEPex1 <- districting %>% 
  filter(SCHOOL_YEAR == "2020") %>% 
  select( NUMBER_CLASSES_3_IEP,
          NUMBER_CLASSES_4_IEP,
          NUMBER_CLASSES_5_IEP,
          NUMBER_CLASSES_6_IEP,
          NUMBER_CLASSES_7_PLUS_IEP,
         riding_code) %>% 
  group_by(riding_code) %>% 
  summarise(`+3 IEPs` = sum(NUMBER_CLASSES_3_IEP,
                            NUMBER_CLASSES_4_IEP,
                            NUMBER_CLASSES_5_IEP,
                            NUMBER_CLASSES_6_IEP,
                            NUMBER_CLASSES_7_PLUS_IEP)) %>%
  drop_na() %>% 
  full_join(BC_data, by = "riding_code")

IEP <- ggplot(IEPex1, aes(x = long, y = lat, group = group, fill = `+3 IEPs`)) +
  geom_polygon() +
  coord_fixed() +
  theme_mapcan() +
  # labs(title = "") +
  scale_fill_continuous(name = "Total Classes with >=3 Students with IEPs")

ggplotly(IEP)
# ggplotly(EL)
# gridExtra::grid.arrange(IEP, EL, ncol = 2)
# plotly::ggplotly(IEP)
# renderPlotly(IEP)

```
### Longitudinal Classes with >3 IEPs

```{r IEP_time}

Plot_Longitudinal1 <- districting %>% 
  mutate(SCHOOL_YEAR = as.numeric(SCHOOL_YEAR),
          totalIEPs = NUMBER_CLASSES_3_IEP,
          NUMBER_CLASSES_4_IEP,
          NUMBER_CLASSES_5_IEP,
          NUMBER_CLASSES_6_IEP,
          NUMBER_CLASSES_7_PLUS_IEP,)

longitudinalBC1 <- ggplot(Plot_Longitudinal1, aes(SCHOOL_YEAR, totalIEPs))+
  geom_line(aes(group = DISTRICT_NAME),
            color = "gray75",
            alpha = 0.3) +
  geom_line(data = filter(Plot_Longitudinal1,
      DISTRICT_NAME == "Surrey"),
           color = "#D95F02") +
  geom_line(data = filter(Plot_Longitudinal1,
      DISTRICT_NAME == "Vancouver"),
      color = "#7570B3") +
  geom_line(data = filter(Plot_Longitudinal1,
      DISTRICT_NAME == "Greater Victoria"),
           color = "#E7298A") +
  theme_minimal() +
  labs(title = "Classrooms with +3 IEPs by School District") +
       xlab("School Year") +
       ylab("IEPs in Overcapacity Classrooms") +
  theme(legend.position = "none") +
    annotate(geom = "text", x = 2012, y = 830, label = "Surrey", hjust = "left") +
    annotate(geom = "text", x = 2008, y = 1080, label = "Vancouver", hjust = "left") +
    annotate(geom = "text", x = 2012, y = 340, label = "Victoria", hjust = "left")

ggplotly(longitudinalBC1)


```



English Learners
=============================

Column {.tabset}
-----------------------------------------------------------------------

### Classes with Minimum Three Students Designated as English Learners

This is another interactive figure which indicates the amount of classrooms in each district that are composed of at least one student who is identified as an English Learner. This is important because the combination of English Learners and students with individualized needs frequently triggers additional supports to be put in place for classroom teachers to make accommodations and modifications for all students who may require additional support.. Future iterations of this figure may demonstrate
this relationship over time for the province of BC.

```{r ELexample}

ELex1 <- districting %>% 
  filter(SCHOOL_YEAR == "2020") %>% 
  select(NUMBER_CLASSES_1_ELL,
         NUMBER_CLASSES_2_ELL,
         NUMBER_CLASSES_3_ELL,
         NUMBER_CLASSES_4_ELL,
         NUMBER_CLASSES_5_ELL,
         NUMBER_CLASSES_6_ELL,
         NUMBER_CLASSES_7_PLUS_ELL,
         riding_code) %>%
  group_by(riding_code) %>%
  summarise(`+1 ELs` = sum( NUMBER_CLASSES_1_ELL, 
                            NUMBER_CLASSES_2_ELL,
                            NUMBER_CLASSES_3_ELL,
                            NUMBER_CLASSES_4_ELL,
                            NUMBER_CLASSES_5_ELL,
                            NUMBER_CLASSES_6_ELL,
                            NUMBER_CLASSES_7_PLUS_ELL)) %>%
  drop_na() %>%
  full_join(BC_data, by = "riding_code") 

EL <- ggplot(ELex1, aes(x = long, y = lat, group = group, fill = `+1 ELs`)) +
  geom_polygon() +
  coord_fixed() +
  theme_mapcan() +
  scale_fill_continuous(name = "Total Students in Classes with >3 ELs")

ggplotly(EL)

```

### Longitudinal Classes with >3 ELs


```{r EL_time}

Plot_Longitudinal <- districting %>% 
  mutate(SCHOOL_YEAR = as.numeric(SCHOOL_YEAR),
          totalELs = NUMBER_CLASSES_3_ELL +
           NUMBER_CLASSES_4_ELL +
           NUMBER_CLASSES_5_ELL +
           NUMBER_CLASSES_6_ELL +
           NUMBER_CLASSES_7_PLUS_ELL)

longitudinalBC <- ggplot(Plot_Longitudinal, aes(SCHOOL_YEAR, totalELs))+
  geom_line(aes(group = DISTRICT_NAME),
            color = "gray75",
            alpha = 0.3) +
  # gghighlight::gghighlight(unhighlighted_params = list(alpha=0.3)) +
  geom_line(data = filter(Plot_Longitudinal,
      DISTRICT_NAME == "Surrey"),
           color = "#D95F02") +
  geom_line(data = filter(Plot_Longitudinal,
      DISTRICT_NAME == "Vancouver"),
      color = "#7570B3") +
  geom_line(data = filter(Plot_Longitudinal,
      DISTRICT_NAME == "Greater Victoria"),
           color = "#E7298A") +
  theme_minimal() +
  labs(title = "Classrooms with +3 ELs by School District") +
       xlab("School Year") +
       ylab("ELs in Overcapacity Classrooms") +
  theme(legend.position = "none") +
    annotate(geom = "text", x = 2017, y = 3250, label = "Surrey", hjust = "left") +
    annotate(geom = "text", x = 2008, y = 3350, label = "Vancouver", hjust = "left") +
    annotate(geom = "text", x = 2012, y = 500, label = "Victoria", hjust = "left")

ggplotly(longitudinalBC)


```

Combined Map
=============================

Column {data-width=350}
-----------------------------------------------------------------------

### EL Juxtaposed IEP


```{r Combined}

gridExtra::grid.arrange(IEP, EL, ncol = 2)

````



